---
title: "Dr. Paul Schmidt"
output: 
  html_document:
    includes:
      in_header: header.html
      after_body: footer.html
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
# formatting tables for html output
options(knitr.kable.NA = '')
pacman::p_load(kableExtra, formattable ,htmltools)
pacman::p_load(tidyverse)
```

```{r import, echo=FALSE, warning=FALSE, message=FALSE}
text <-  read.table(file = "https://raw.githubusercontent.com/SchmidtPaul/CV/master/contactinfo/infotext.csv", sep = ";", header = T, encoding = "UTF-8") %>%
  rename(English = 1) %>% 
  tibble

workshops <- read.table(file = "https://raw.githubusercontent.com/SchmidtPaul/CV/master/contactinfo/workshophistory.csv", sep = ";", header = T, encoding = "UTF-8") %>%
  rename(Time = 1) %>% tibble
```

<!-- hyperlink style -->
<style>
a:link {
  color: #003f75ff;
  background-color: transparent;
  text-decoration: none;
}

a:visited {
  color: #003f75ff;
  background-color: transparent;
  text-decoration: none;
}

a:hover {
  color: #003f75ff;
  background-color: transparent;
  text-decoration: underline;
}

a:active {
  color: #003f75ff;
  background-color: transparent;
  text-decoration: underline;
}
</style> 

<!-- Image -->
<style>
img {border-radius: 50%;}
</style>
<img src="https://avatars3.githubusercontent.com/u/31928213?s=400&u=d5ca44cc7f39c7f4e2e79b4cffeeb3d0d0a321fb&v=4" style="width:33%; margin-left: 20px" align="right" alt="Avatar">

<!-- Contact symbols -->
<style>
p.small {line-height: 1;}
</style>

<p class="small">
<i class="fa fa-home" style="color: #003f75ff;">
</i> Hamburg, Germany </br>

Contact </br>
<a href="mailto:schmidtpaul1989@outlook.com">
<i class="fa fa-envelope" style="color: #003f75ff;">
</i> schmidtpaul1989@outlook.com </a> </br>
<a href="https://github.com/SchmidtPaul/">
<i class="fa fa-github" style="color: #003f75ff;">
</i> GitHub </a> </br>
<a href="https://www.linkedin.com/in/schmidtpaul1989/">
<i class="fa fa-linkedin" style="color: #003f75ff;">
</i> LinkedIn </a> </br>
<a href="https://www.researchgate.net/profile/Paul_Schmidt17">
<i class="fa fa-pencil" style="color: #003f75ff;">
</i> ResearchGate </a> </br>
<a href="https://www.xing.com/profile/Paul_Schmidt393/cv">
<i class="fa fa-xing" style="color: #003f75ff;">
</i> Xing </a> 

Websites </br>
<a href="https://schmidtpaul.github.io/crashcouRse/">
<i class="fa fa-window-maximize" style="color: #003f75ff;">
</i> CrashcouRse </a> </br>
<a href="https://schmidtpaul.github.io/DSFAIR/">
<i class="fa fa-window-maximize" style="color: #003f75ff;">
</i> DSFAIR: Data Science for Agriculture in R </a> </br>
<a href="https://schmidtpaul.github.io/MMFAIR/">
<i class="fa fa-window-maximize" style="color: #003f75ff;">
</i> MMFAIR: Mixed Models for Agriculture in R </a> </br>

</p>

```{r text, echo=FALSE, warning=FALSE, message=FALSE}
text %>% 
  kbl() %>%
  kable_styling() %>% 
  row_spec(0, color = "lightgrey")
```

```{r workshops, echo=FALSE, warning=FALSE, message=FALSE}
GerFlagURL <- "https://cdn.countryflags.com/thumbs/germany/flag-round-250.png"
USAFlagURL <- "https://cdn.countryflags.com/thumbs/united-states-of-america/flag-round-250.png"

LangFlags <- workshops %>% 
  select(Language) %>% 
  mutate(LangURL = case_when(Language == "Ger" ~ GerFlagURL,
                             Language == "Eng" ~ USAFlagURL,
                             T ~ NA_character_)) %>% 
  pull(LangURL)

workshops %>% 
  select(-URL) %>% 
  kbl() %>%
  kable_minimal(full_width = T, font_size = 10, 
                bootstrap_options = c("hover", "condensed")) %>% 
  column_spec(3, image = spec_image(LangFlags, 50, 50))
```
