---
title: "Heterogeneous error variance"
output: 
  html_document:
    includes:
      in_header: header.html
      after_body: footer.html	
params:
  hilang:
    - sas					 
---

```{r, echo=FALSE, message=FALSE, warning=FALSE, purl=FALSE}
# markdown options
options(knitr.kable.NA = '') # show NA as empty cells in output tables
pacman::p_load(kableExtra, formattable, htmltools) # packages for better formatting tables for html output
```

```{r, child="_hilang_setup.Rmd", purl=FALSE}
```

```{r, message=F, warning=F, error=F}
# packages
pacman::p_load(conflicted, # package function conflicts
               dplyr, purrr, stringr, tibble, tidyr, # data handling
               nlme, lme4, glmmTMB, sommer, # mixed modelling
               AICcmodavg, broom.mixed)     # mixed model extractions

# package function conflicts
conflict_prefer("filter", "dplyr")

# data
dat <- agridat::mcconway.turnip %>%
  as_tibble() %>%
  mutate(densf = density %>% as.factor)
```

```{r, echo=FALSE, purl=FALSE}
dat %>%
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE) %>% 
  scroll_box(height = "200px")
```

<br/>

# Motivation
This is a randomized complete block design (4 blocks) with three treatment factors: genotype, date and density, leading to 16 treatment level combinations (2 genotypes, 2 planting dates, 4 densities) [(Piepho, 2009)](https://acsess.onlinelibrary.wiley.com/doi/abs/10.2134/agronj2008.0226x){target="_blank"}. It can be argued that heterogeneous error variances (*i.e.* heteroscedasticity) for two of the treatments should be considered:

<div class = "row"> <div class = "col-md-6">
```{r, purl=FALSE}
boxplot(yield ~ date, data = dat)
```
</div> <div class = "col-md-6">
```{r, purl=FALSE}
boxplot(yield ~ density, data = dat)
```
</div> </div>

# 5 Models {.tabset .tabset-fade .tabset-pills}
We therefore set up 5 models, which only differ in the variance structure in the error term. More specifically, we allow for different heterogeneous error variances / heteroscedascity per group for the effect levels of Date and Density. This type of variance structure is sometimes also referred to as [*diagonal*](variance_structures.html#diagonal){target="_blank"}. 

Regarding the effects in the model, we take the same approach as [Piepho (2009)](https://acsess.onlinelibrary.wiley.com/doi/abs/10.2134/agronj2008.0226x){target="_blank"}, which calls for fixed main effects for `gen`, `date` and `densf` as well as all their interaction effects, plus random `block` effects.

The different variance structures in the error term of the 5 models are summarised in the table below (only mods 1-4 can be found in [Piepho (2009)](https://acsess.onlinelibrary.wiley.com/doi/abs/10.2134/agronj2008.0226x){target="_blank"}). Further notice the difference between *mod4* and *mod5*: *mod4* includes the direct product (*a.k.a.* Kronecker product) structure of the two diag-structures of *Date* and *Density*. *mod5* simply includes a diag-structure for the *Date*-*Density*-combinations (see our [summary on variance structures](variance_structures.html){target="_blank"} for more info).

```{r, echo=FALSE, purl=FALSE}
MODTAB <- tibble(Model = paste0("mod", 1:5) %>% cell_spec(bold=T),
       Block    = "Identity",
       Genotype = "Identity",
       Date     = c("Identity", "Identity", "Diagonal", "Diagonal", "Diag-"),
       Density  = c("Identity", "Diagonal", "Identity", "Diagonal", "onal"),
       parameters = c(1,2,4,5,8),
       estimates  = c(1,2,4,8,8)) %>% 
  mutate(Date    = ifelse(Date    %in% c("Diagonal", "Diag-"), cell_spec(Date,    bold=T), Date),
         Density = ifelse(Density %in% c("Diagonal", "onal"),  cell_spec(Density, bold=T), Density))

names(MODTAB)[6] <- paste0(names(MODTAB)[6], footnote_marker_alphabet(1))
names(MODTAB)[7] <- paste0(names(MODTAB)[7], footnote_marker_alphabet(1))

MODTAB %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE) %>% 
  add_header_above(c(" ", 
                     "Term in multiplicative variance structure"=4, 
                     "Number of variance"=2)) %>% 
  footnote(alphabet = "ignoring the random block effects",
           footnote_as_chunk = T)
```

## nlme
To obtain heterogeneous variances in `nlme`, we need to use the *variance function* [`varIdent()`](https://cran.r-project.org/web/packages/nlme/nlme.pdf#Rfn.varIdent.1){target="_blank"} in the [`weights=`](https://cran.r-project.org/web/packages/nlme/nlme.pdf#Rfn.lme.1){target="_blank"} argument, which is *used to allow for different variances according to the levels of a classification factor*. For the multiplicative variance structure in *mod4*, we can combine two variance functions via [`varComb()`](https://cran.r-project.org/web/packages/nlme/nlme.pdf#Rfn.varComb.1){target="_blank"}. Since it is not possible to pass an interaction term to the `varIdent()` function as `varIdent(form=~1|date:densf)`, we must manually create a column that combines the two columns in order to fit *mod5*:

```{r}
dat <- dat %>%
  mutate(date_densf = interaction(date, densf)) # needed for mod5

mod1.nlme <- nlme::lme(fixed   = yield ~ gen * date * densf,  
                       random  = ~ 1 | block, 
                       weights = NULL, # default, i.e. homoscedastic errors
                       data    = dat)

mod2.nlme <- mod1.nlme %>%
  update(weights = varIdent(form =  ~ 1 | date))

mod3.nlme <- mod1.nlme %>%
  update(weights = varIdent(form =  ~ 1 | densf))

mod4.nlme <- mod1.nlme %>%
  update(weights = varComb(varIdent(form =  ~ 1 | date),
                           varIdent(form =  ~ 1 | densf))) 

mod5.nlme <- mod1.nlme %>%
  update(weights = varIdent(form =  ~ 1 | date_densf))
```

## lme4
The short answer here is that with `lme4` it is **not possible** to fit any variance structures, so that in this chapter only `mod1` could be modeled: 

```{r}
mod1.lme4 <- lmer(formula = yield ~ gen * date * densf + (1 | block),
                  data    = dat)

# mod2.lme4 - not possible

# mod3.lme4 - not possible

# mod4.lme4 - not possible

# mod5.lme4 - not possible
```

More specifically, we can read in an [`lme4` vigniette](https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf){target="_blank"}: *"The main advantage of `nlme` relative to `lme4` is a user interface for fitting models with structure in the residuals (various forms of heteroscedasticity and autocorrelation) and in the random-effects covariance matrices (e.g., compound symmetric models).  With some extra effort, the computational machinery of `lme4` can be used to fit structured models that the basic `lmer` function cannot handle (see [Appendix A](https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf#%5B%7B%22num%22%3A15%2C%22gen%22%3A0%7D%2C%7B%22name%22%3A%22XYZ%22%7D%2C81%2C733.028%2Cnull%5D){target="_blank"})"*

Michael Clark [puts it as](https://m-clark.github.io/mixed-models-with-R/extensions.html#heterogeneous-variance){target="_blank"} *"Unfortunately, lme4 does not provide the ability to model the residual covariance structure, at least [not in a straightforward fashion](https://bbolker.github.io/mixedmodels-misc/notes/corr_braindump.html){target="_blank"}"*

**Thus, there is no info on this package for this chapter beyond this point, except for model 1.**

## glmmTMB
In `glmmTMB()` it is -to our knowledge- not possible to adjust the variance structure of the error.

>Just like in `nlme`, there is a `weights=` argument in `glmmTMB()`. However, to our understanding, they have different functions: 
> 
>In `nlme`, it requires *"an optional `varFunc` object or one-sided formula describing the within-group heteroscedasticity structure"* [(nlme RefMan)](https://cran.r-project.org/web/packages/glmmTMB/glmmTMB.pdf#Rfn.glmmTMB.1){target="_blank"} and we make use of this in the chapter at hand.   
>
>In `glmmTMB`, the [RefMan](https://cran.r-project.org/web/packages/glmmTMB/glmmTMB.pdf#Rfn.glmmTMB.1){target="_blank"} only states *"weights, as in `glm`. Not automatically scaled to have sum 1"*. Following this trail, the [`glm` documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/glm.html){target="_blank"} description for the `weights=` argument is *"an optional vector of ‘prior weights’ to be used in the fitting process. Should be NULL or a numeric vector."*.
> Accordingly, it cannot be used to allow for heterogeneous error variances in this package.

We can, however, "*fix the residual variance  to  be  0  (actually  a  small  non-zero  value)*" and therefore "*force variance into the random effects*" [(glmmTMB RefMan)](https://cran.r-project.org/web/packages/glmmTMB/glmmTMB.pdf#Rfn.glmmTMB.1){target="_blank"} via adding the `dispformula = ~ 0` argument. Thus, when doing so, we need to make sure to also add a random term to the model with the desired variance structure. By taking both of these actions, we are essentially mimicing the error (variance) as a random effect (variance). We achieve this by first creating a `unit` column in the data with different entries for each data point:

```{r}
dat <- dat %>%
  mutate(unit = 1:n() %>% as.factor) # new column with running number
```

We should now be able to mimic the error variance via the random term `(1|unit)`. We can verify this quickly, by modelling *mod1* with a homoscedastic error variance in the standard fashion (`mod1`) and via a mimicked error term (`mod1b`). For all other models we make use of the `diag()` function. Notice that we must write `diag(TERM + 0|unit)`, since leaving out the `+ 0` would by default lead to estimating not only the desired heterogeneous variances, but an additional overall variance.

Notice that [at the moment, it is not possible](https://github.com/glmmTMB/glmmTMB/issues/592){target="_blank"}  to fit `mod4` with `glmmTMB`.

```{r, warning=FALSE}
mod1.glmm <- glmmTMB(formula = yield ~ gen * date * densf + (1 | block), 
                     dispformula = ~ 1,  # = default i.e. homoscedastic error variance
                     REML = TRUE,        # needs to be stated since default = ML
                     data = dat) 

mod1b.glmm <- mod1.glmm %>%
  update(. ~ . + (1 | unit), # add random term to mimic homoscedastic error variance
         dispformula = ~ 0) # fix original error variance to 0

mod2.glmm <- mod1.glmm %>% 
  update(. ~ . + diag(date + 0 | unit), dispformula = ~ 0)

mod3.glmm <- mod1.glmm %>% 
  update(. ~ . + diag(densf + 0 | unit), dispformula = ~ 0)

# mod4/multiplicative variance structure not possible.

mod5.glmm <- mod1.glmm %>% 
  update(. ~ . + diag(date:densf + 0 | unit), dispformula = ~ 0)
```

As can be seen below, mimicing the error (variance) worked, as it leads to comparable variance component estimates:

<div class = "row"> <div class = "col-md-6">
```{r}
mod1.glmm %>% VarCorr()
```
</div> <div class = "col-md-6">
```{r}
mod1b.glmm %>% VarCorr()
```
</div> </div>

## sommer
To allow for different variance structures in the error term in `sommer`, we need to use the [`vs()`](https://cran.r-project.org/web/packages/sommer/sommer.pdf#Rfn.vs.1){target="_blank"} function in the [`rcov=`](https://cran.r-project.org/web/packages/sommer/sommer.pdf#Rfn.mmer.1){target="_blank"} argument. More specifically, we need the [`ds()`](https://cran.r-project.org/web/packages/sommer/sommer.pdf#Rfn.ds.1){target="_blank"} function to obtain heterogeneous variances.

```{r, message=FALSE, warning=FALSE}
mod1.somm <- mmer(fixed  = yield ~ gen * date * densf, 
                  random = ~ block,
                  rcov   = ~ units, # default
                  data   = dat, verbose=F)

mod2.somm <- mmer(fixed  = yield ~ gen * date * densf, 
                  random = ~ block,
                  rcov   = ~ vs(ds(date), units), 
                  data   = dat, verbose=F)

mod3.somm <- mmer(fixed  = yield ~ gen * date * densf, 
                  random = ~ block,
                  rcov   = ~ vs(ds(densf), units), 
                  data   = dat, verbose=F)

mod4.somm <- mmer(fixed  = yield ~ gen * date * densf, 
                  random = ~ block,
                  rcov   = ~ vs(ds(date), units) + vs(ds(densf), units), 
                  data   = dat, verbose=F)

mod5.somm <- mmer(fixed  = yield ~ gen * date * densf, 
                  random = ~ block,
                  rcov   = ~ vs(ds(date), ds(densf), units), 
                  data   = dat, verbose=F)
```

## SAS

```{r, eval=FALSE, hilang="sas", purl=FALSE}
/* mod1 */
proc mixed data=dat;
class gen date densf block;
model yield = gen * date * densf;
random block;
ods output covparms=mod1sasVC FitStatistics=mod1sasAIC;
run;

/* mod2 */
proc mixed data=dat;
class gen date densf block;
model yield = gen * date * densf;
random block;
repeated / group=date;
ods output covparms=mod2sasVC FitStatistics=mod2sasAIC;
run;

/* mod3 */
proc mixed data=dat;
class gen date densf block;
model yield = gen * date * densf;
random block;
repeated / group=densf;
ods output covparms=mod3sasVC FitStatistics=mod3sasAIC;
run;

/* mod4 */
proc mixed data=dat;
class gen date densf block;
model yield = gen * date * densf;
random block;
repeated / local=exp(date densf); 
ods output covparms=mod4sasVC FitStatistics=mod4sasAIC;
run;

/* mod5 */
proc mixed data=dat;
class gen date densf block;
model yield = gen * date * densf;
random block;
repeated / group=date*densf;
ods output covparms=mod5sasVC FitStatistics=mod5sasAIC;
run;
```

<br/>

# Variance Component Extraction {.tabset .tabset-fade .tabset-pills}

After fitting the models, we would now like to extract the variance component estimates.

## nlme
As far as we know, it is quite cumbersome to extract variance component estimates from `nlme` objects (in a table format we are used to), even with helper packages such as `broom.mixed` (see [github issue](https://github.com/bbolker/broom.mixed/issues/96){target="_blank"} ). Beneath is the the most elegant approach we have so far when trying to get similarly formatted output for all 5 models. The general approach is to first get a table with a column for 

+ all levels / level-combinations for which heterogeneous error variances were fit
+ their respective value of the [model-object's `varStruct`](https://cran.r-project.org/web/packages/nlme/nlme.pdf#Rfn.nlmeStruct.1){target="_blank"} and
+ the overall value of the [model-object's `sigma`](https://cran.r-project.org/web/packages/nlme/nlme.pdf#Rfn.nlmeObject.1){target="_blank"} (= standard deviation for error term).



By multiplying the latter two, we receive the respective standard error estimate, which needs to be squared in order to obtain the respective variance component estimate.

**mod1**
```{r}
mod1.nlme.VC <- tibble(grp = "homoscedastic", varStruct = 1) %>% 
  mutate(sigma         = mod1.nlme$sigma) %>% 
  mutate(StandardError = sigma*varStruct) %>% 
  mutate(Variance      = StandardError^2)
```
```{r, echo=FALSE, purl=FALSE}
mod1.nlme.VC %>%
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod2**
```{r}
mod2.nlme.VC <- mod2.nlme$modelStruct$varStruct %>%
  coef(unconstrained = FALSE, allCoef = TRUE) %>%
  enframe(name = "grp", value = "varStruct") %>%
  mutate(sigma         = mod2.nlme$sigma) %>%
  mutate(StandardError = sigma * varStruct) %>%
  mutate(Variance      = StandardError ^ 2)
```
```{r, echo=FALSE, purl=FALSE}
mod2.nlme.VC %>%
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod3**
```{r}
mod3.nlme.VC <- mod3.nlme$modelStruct$varStruct %>%
  coef(unconstrained = FALSE, allCoef = TRUE) %>%
  enframe(name = "grp", value = "varStruct") %>%
  mutate(sigma         = mod3.nlme$sigma) %>%
  mutate(StandardError = sigma * varStruct) %>%
  mutate(Variance      = StandardError ^ 2)
```
```{r, echo=FALSE, purl=FALSE}
mod3.nlme.VC %>%
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod4**
```{r}
mod4.nlme.varStruct.A <- mod4.nlme$modelStruct$varStruct$A %>%
  coef(unconstrained = FALSE, allCoef = TRUE) %>%
  enframe(name = "grpA", value = "varStructA")

mod4.nlme.varStruct.B <- mod4.nlme$modelStruct$varStruct$B %>%
  coef(unconstrained = FALSE, allCoef = TRUE) %>%
  enframe(name = "grpB", value = "varStructB")

mod4.nlme.VC <- expand.grid(mod4.nlme.varStruct.A$grpA,
                            mod4.nlme.varStruct.B$grpB,
                            stringsAsFactors = FALSE) %>%
  rename(grpA = Var1, grpB = Var2) %>%
  left_join(x = ., y = mod4.nlme.varStruct.A, by = "grpA") %>%
  left_join(x = ., y = mod4.nlme.varStruct.B, by = "grpB") %>%
  mutate(sigma         = mod4.nlme$sigma) %>%
  mutate(StandardError = sigma * varStructA * varStructB) %>%
  mutate(Variance      = StandardError ^ 2)
```
```{r, echo=FALSE, purl=FALSE}
mod4.nlme.VC %>%
  arrange(grpA, grpB) %>% 
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod5**
```{r}
mod5.nlme.VC <- mod5.nlme$modelStruct$varStruct %>%
  coef(unconstrained = FALSE, allCoef = TRUE) %>%
  enframe(name = "grp", value = "varStruct") %>%
  separate(grp, sep = "[.]", into = c("grpA", "grpB")) %>%
  mutate(sigma         = mod5.nlme$sigma) %>%
  mutate(StandardError = sigma * varStruct) %>%
  mutate(Variance      = StandardError ^ 2)
```
```{r, echo=FALSE, purl=FALSE}
mod5.nlme.VC %>%
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

> We want to point out that the `tidy(effects="ran_pars")` function of the  `broom.mixed` package (which we used it in this chapter to obtain the variance component estimates from `glmmTMB()` models) [actually works](https://cran.r-project.org/web/packages/broom.mixed/vignettes/broom_mixed_intro.html){target="_blank"} on `lme()` objects as well (but it does not on `gls()` and `nlme()`). We did not use it here, however, since it only seems to provide the sigma estimate and not the varStruct estimates (see [github issue](https://github.com/bbolker/broom.mixed/issues/96){target="_blank"}).

## lme4
In order to extract variance component estimates from `lme4` objects (in a table format we are used to), we here use the [`tidy()`](https://cran.r-project.org/web/packages/broom.mixed/broom.mixed.pdf#Rfn.glmmTMB.Rul.tidiers.1){target="_blank"} function from the helper package [`broom.mixed`]( https://CRAN.R-project.org/package=broom.mixed){target="_blank"}:

**mod1**
```{r}
mod1.lme4.VC <- mod1.lme4 %>%
  tidy(effects = "ran_pars", scales = "vcov") %>%
  separate(term, sep = "__", into = c("term", "grp")) %>%
  filter(group == "Residual")
```
```{r, echo=FALSE, purl=FALSE}
mod1.lme4.VC %>%
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

## glmmTMB

In order to extract variance component estimates from `glmmTMB` objects (in a table format we are used to), we here use the [`tidy()`](https://cran.r-project.org/web/packages/broom.mixed/broom.mixed.pdf#Rfn.glmmTMB.Rul.tidiers.1){target="_blank"} function from the helper package [`broom.mixed`]( https://CRAN.R-project.org/package=broom.mixed){target="_blank"}:

**mod1**
```{r}
mod1.glmm.VC <- mod1.glmm %>%
  tidy(effects = "ran_pars", scales = "vcov") %>%
  separate(term, sep = "__", into = c("term", "grp")) %>%
  filter(group == "Residual")
```
```{r, echo=FALSE, purl=FALSE}
mod1.glmm.VC %>%
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod2**
```{r}
mod2.glmm.VC <- mod2.glmm %>%
  tidy(effects = "ran_pars", scales = "vcov") %>%
  separate(term, sep = "__", into = c("term", "grp")) %>%
  filter(group == "unit", term == "var")
```
```{r, echo=FALSE, purl=FALSE}
mod2.glmm.VC %>%
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod3**
```{r}
mod3.glmm.VC <- mod3.glmm %>%
  tidy(effects = "ran_pars", scales = "vcov") %>%
  separate(term, sep = "__", into = c("term", "grp")) %>%
  filter(group == "unit", term == "var")
```
```{r, echo=FALSE, purl=FALSE}
mod3.glmm.VC %>%
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod4**

mod4/multiplicative variance structure not possible.

**mod5**
```{r}
mod5.glmm.VC <- mod5.glmm %>%
  tidy(effects = "ran_pars", scales = "vcov") %>%
  separate(term, sep = "__", into = c("term", "grp")) %>%
  filter(group == "unit", term == "var") %>%
  separate(grp, sep = ":", into = c("grpA", "grpB"))
```
```{r, echo=FALSE, purl=FALSE}
mod5.glmm.VC %>%
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

## sommer
It is quite easy to extract variance component estimates from `sommer` objects in a table format we are used to:

**mod1**
```{r}
mod1.somm.VC <- summary(mod1.somm)$varcomp 
```
```{r, echo=FALSE}
mod1.somm.VC <- mod1.somm.VC %>% 
  as_tibble(rownames="grp") %>% 
  mutate(grp = str_replace(grp, "\\..*", "")) %>% 
  filter(grp!="block")
```
```{r, echo=FALSE, purl=FALSE}
mod1.somm.VC %>% 
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod2**
```{r}
mod2.somm.VC <- summary(mod2.somm)$varcomp
```
```{r, echo=FALSE}
mod2.somm.VC <- mod2.somm.VC %>% 
  as_tibble(rownames="grp") %>% 
  mutate(grp = str_replace(grp, "\\..*", "")) %>% 
  filter(grp!="block") 
```
```{r, echo=FALSE, purl=FALSE}
mod2.somm.VC %>% 
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod3**
```{r}
mod3.somm.VC <- summary(mod3.somm)$varcomp
```
```{r, echo=FALSE}
mod3.somm.VC <- mod3.somm.VC %>% 
  as_tibble(rownames="grp") %>% 
  mutate(grp = str_replace(grp, "\\..*", "")) %>% 
  filter(grp!="block") 
```
```{r, echo=FALSE, purl=FALSE}
mod3.somm.VC %>% 
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod4**

mod4/multiplicative variance structure possible? in progress.

**mod5**
```{r}
mod5.somm.VC <- summary(mod5.somm)$varcomp
```
```{r, echo=FALSE}
mod5.somm.VC <- mod5.somm.VC %>% 
  as_tibble(rownames="grp") %>% 
  mutate(grp = str_replace(grp, "\\.yield-yield", "")) %>% 
  filter(grp!="block") 
```
```{r, echo=FALSE, purl=FALSE}
mod5.somm.VC %>% 
  arrange(grp) %>%
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

## SAS

Thanks to the [`ODS` (Output Delivery System)](https://support.sas.com/rnd/base/ods/){target="_blank"} in SAS, there are no extra steps required to extract the variance component estimates. This was already achieved via the `ods output covparms=modsasVC;` lines in the `PROC MIXED` statements in the previous section. It should be noted, however, that the estimates for `mod4` **do** need further formatting in order to obtain the here presented results. These steps are somewhat similar to those for the `nlme.VC`.

```{r, echo=FALSE, message=FALSE}
sasvc <- readr::read_delim("SAS/heterogeneous_error_variance_results_VC.txt", delim="\t") %>% 
  filter(CovParm != "block")

mod1.sas.VC <- sasvc %>%
  filter(mod==1) %>% 
  mutate(Group="homoscedastic") %>% 
  dplyr::select(CovParm, Group, Estimate)

mod2.sas.VC <- sasvc %>%
  filter(mod==2) %>% 
  # mutate(Group = str_remove(Group, "date ")) %>% 
  dplyr::select(CovParm, Group, Estimate)

mod3.sas.VC <- sasvc %>%
  filter(mod==3) %>% 
  # mutate(Group = str_remove(Group, "densf ")) %>% 
  dplyr::select(CovParm, Group, Estimate)

mod4.sas.VC <- sasvc %>%
  filter(mod==4) %>% 
  dplyr::select(CovParm, Estimate) %>% 
  bind_rows(tibble(CovParm=c("EXP date", "EXP densf"), Estimate=c(0.5581,sum(-1.29,-0.507,1.06)*(-1)))) %>% 
  bind_cols(Group=c("21Aug1990","1","2","4",NA,"28Aug1990","8")) %>% 
  mutate(CovParm = str_remove(CovParm, "EXP "))
mod4.sas.varStruct.A <- mod4.sas.VC %>% filter(CovParm=="date")
mod4.sas.varStruct.B <- mod4.sas.VC %>% filter(CovParm=="densf")
mod4.sas.VC <- expand.grid(mod4.sas.varStruct.A$Group,
                           mod4.sas.varStruct.B$Group,
                           stringsAsFactors = FALSE) %>%
  rename(grpA = Var1, grpB = Var2) %>%
  left_join(x = ., y = mod4.sas.VC %>% dplyr::select(-CovParm) %>% rename(EXPdate=Estimate), 
            by = c("grpA"="Group")) %>%
  left_join(x = ., y = mod4.sas.VC %>% dplyr::select(-CovParm) %>% rename(EXPdensf=Estimate), 
            by = c("grpB"="Group")) %>%
  mutate(sigma    = mod4.sas.VC %>% filter(CovParm=="Residual") %>% pull(Estimate)) %>%
  mutate(Variance = sigma * exp(EXPdate) * exp(EXPdensf)) %>% 
  arrange(grpB, grpA)
  
mod5.sas.VC <- expand.grid(mod4.sas.varStruct.A$Group,
                           mod4.sas.varStruct.B$Group,
                           stringsAsFactors = FALSE) %>% 
  rename(grpA = Var1, grpB = Var2) %>% 
  arrange(grpA, grpB) %>% 
  bind_cols(sasvc %>% filter(mod==5) %>% dplyr::select(CovParm, Estimate)) %>% 
  dplyr::select(CovParm, grpA, grpB, Estimate)
```

**mod1**
```{r, echo=FALSE, purl=FALSE}
mod1.sas.VC %>% 
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod2**
```{r, echo=FALSE, purl=FALSE}
mod2.sas.VC %>% 
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod3**
```{r, echo=FALSE, purl=FALSE}
mod3.sas.VC %>% 
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod4**
```{r, echo=FALSE, purl=FALSE}
mod4.sas.VC %>% 
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod5**
```{r, echo=FALSE, purl=FALSE}
mod5.sas.VC %>% 
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

# Model Fit {.tabset .tabset-fade .tabset-pills}
As is standard procedure, we can do a model selection based on goodness-of-fit statistics such as the [AIC](https://www.wikiwand.com/en/Akaike_information_criterion#){target="_blank"} (Akaike information criterion). In order to have a direct comparison to Table 1 in [(Piepho, 2009)](https://acsess.onlinelibrary.wiley.com/doi/abs/10.2134/agronj2008.0226x){target="_blank"}, we also calculated the deviance (*i.e.* -2*loglikelihood) for each model.

For `nlme`, `lme4` and `glmmTMB`, we make use of the [`aictab()` function](https://cran.r-project.org/web/packages/AICcmodavg/AICcmodavg.pdf#Rfn.aictab.1){target="_blank"} from the helper package [AICcmodavg](https://CRAN.R-project.org/package=AICcmodavg){target="_blank"}.

## nlme

```{r, warning=FALSE}
AIC.nlme <- aictab(list(mod1.nlme, mod2.nlme, mod3.nlme, mod4.nlme, mod5.nlme)) %>%
  mutate(Deviance = -2 * Res.LL) # compute deviance manually
```

```{r, echo=FALSE, purl=FALSE}
AIC.nlme %>%  
  dplyr::select(Modnames, K, AICc, Delta_AICc, Res.LL, Deviance) %>% 
  mutate_at(vars(AICc:Deviance), round, 1) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

## lme4
```{r, warning=FALSE}
AIC.lme4 <- aictab(list(mod1.lme4)) %>% # Mods 2-5 are missing
  mutate(Deviance = -2 * Res.LL) # compute deviance manually
```
```{r, echo=FALSE, purl=FALSE}
AIC.lme4 %>%  
  dplyr::select(Modnames, K, AICc, Delta_AICc, Res.LL, Deviance) %>% 
  mutate_at(vars(AICc:Deviance), round, 1) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

## glmmTMB
```{r, warning=FALSE}
AIC.glmm <- aictab(list(mod1.glmm, mod2.glmm, mod3.glmm, mod5.glmm), 
                   modnames=c("Mod1","Mod2","Mod3","Mod5")) %>% # Mod4 is missing
  mutate(Deviance = -2*LL) # compute deviance manually
```

```{r, echo=FALSE, purl=FALSE}
AIC.glmm %>%  
  dplyr::select(Modnames, K, AICc, Delta_AICc, LL, Deviance) %>%  
  mutate_at(vars(AICc:Deviance), ~round(., 1)) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

## sommer

Unfortunately, the `aictab()` function does not work on `mmer` objects. We therefore create a comparable table manually. Notice further that in the `sommer` package, the likelihood is calculated in a different way compared to `nlme`, `lme4` and `glmmTMB`, leading to deviating values between packages.

```{r}
somm.mods <- list(mod1.somm, mod2.somm, mod3.somm, mod5.somm)

AIC.somm <- tibble(
    Modnames = paste0("Mod", c(1:3, 5)), # Mod4 is missing
    AIC = somm.mods %>% map("AIC") %>% unlist,
    LL  = somm.mods %>% map("monitor") %>% lapply(. %>% `[`(1, ncol(.))) %>% unlist) %>% # last element in first row of "monitor"
  mutate(Deviance = -2 * LL) %>% # compute deviance manually
  arrange(AIC)
```
```{r, echo=FALSE, purl=FALSE}
AIC.somm %>%  
  mutate_at(vars(AIC:Deviance), ~round(., 1)) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

## SAS

Thanks to the [`ODS` (Output Delivery System)](https://support.sas.com/rnd/base/ods/){target="_blank"} in SAS, there are no extra steps required to extract the model fit statistics. This was already achieved via the `ods output FitStatistics=modsasAIC;` lines in the `PROC MIXED` statements in the previous section.

```{r, echo=FALSE, message=FALSE}
AIC.sas <- readr::read_delim("SAS/heterogeneous_error_variance_results_AIC.txt", delim="\t") %>% 
  mutate(Descr=str_remove(Descr, " \\(kleiner ist besser\\)"),
         Modnames=paste0("Mod",mod)) %>% 
  dplyr::select(-mod) %>% 
  pivot_wider(names_from="Descr", values_from="Value") %>% 
  rename(Deviance = `-2 Res Log-Likelihood`)
```
```{r, echo=FALSE, purl=FALSE}
AIC.sas %>%  
  mutate_at(vars(Deviance:BIC), ~round(., 1)) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

# Summary {.tabset .tabset-fade}

## Syntax

package | model syntax
---|---
`nlme` | `weights = varIdent(form = ~ 1 | TERM)`
`lme4` | <span style="color:red">not possible</span>
`glmmTMB` | `random = ~ diag(TERM + 0 | unit), dispformula =  ~ 0`
`sommer` | `rcov = ~ vs(ds(TERM), units)`
`SAS` | `repeated / group=TERM;`

## VarComp

**mod1**
```{r, echo=FALSE}
plyr::join_all(list(mod1.nlme.VC %>% dplyr::select(grp, Variance) %>% rename(nlme=Variance),
                    mod1.lme4.VC %>% mutate(grp="homoscedastic") %>% 
                      dplyr::select(grp, estimate) %>% rename(lme4=estimate),
                    mod1.glmm.VC %>% mutate(grp="homoscedastic") %>% 
                      dplyr::select(grp, estimate) %>% rename(glmmTMB=estimate),
                    mod1.somm.VC %>% mutate(grp="homoscedastic") %>% 
                      dplyr::select(grp, VarComp) %>% rename(sommer=VarComp),
                    mod1.sas.VC %>% mutate(grp="homoscedastic") %>% 
                      dplyr::select(grp, Estimate) %>% rename(SAS=Estimate)
                    ), 
               by="grp", type="left") %>%
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod2**
```{r, echo=FALSE}
plyr::join_all(list(mod2.nlme.VC %>% dplyr::select(grp, Variance) %>% rename(nlme=Variance),
                    mod2.nlme.VC %>% dplyr::select(grp) %>% mutate(lme4=NA),
                    mod2.glmm.VC %>% dplyr::select(grp, estimate) %>% 
                      mutate(grp = str_remove(grp, "date")) %>% rename(glmmTMB=estimate),
                    mod2.somm.VC %>% dplyr::select(grp, VarComp) %>% 
                      mutate(grp = str_replace(grp, ":units", "")) %>% rename(sommer=VarComp),
                    mod2.sas.VC %>% mutate(grp = str_remove(Group, "date ") %>% 
                                                 str_replace("UG","ug")) %>% 
                      dplyr::select(grp, Estimate) %>% rename(SAS=Estimate)), 
               by="grp", type="left") %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod3**
```{r, echo=FALSE}
plyr::join_all(list(mod3.nlme.VC %>% dplyr::select(grp, Variance) %>% rename(nlme=Variance),
                    mod3.nlme.VC %>% dplyr::select(grp) %>% mutate(lme4=NA),
                    mod3.glmm.VC %>% dplyr::select(grp, estimate) %>% 
                      mutate(grp = str_remove(grp, "densf")) %>% rename(glmmTMB=estimate),
                    mod3.somm.VC %>% dplyr::select(grp, VarComp) %>% 
                      mutate(grp = str_replace(grp, ":units", "")) %>% rename(sommer=VarComp),
                    mod3.sas.VC %>% 
                      mutate(grp = str_remove(Group, "densf ")) %>% 
                      dplyr::select(grp, Estimate) %>% rename(SAS=Estimate)
                    ), 
               by="grp", type="left") %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod4**
```{r, echo=FALSE}
plyr::join_all(list(mod4.nlme.VC %>% dplyr::select(grpA, grpB, Variance) %>% rename(nlme=Variance),
                    mod4.nlme.VC %>% dplyr::select(grpA, grpB) %>% mutate(lme4=NA),
                    mod4.nlme.VC %>% dplyr::select(grpA, grpB) %>% mutate(glmmTMB=NA),
                    mod4.nlme.VC %>% dplyr::select(grpA, grpB) %>% mutate(sommer="in progress"),
                    mod4.sas.VC %>% dplyr::select(grpA, grpB, Variance) %>% rename(SAS=Variance)), 
               by=c("grpA", "grpB"), type="left") %>% 
  arrange(grpA, grpB) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

**mod5**
```{r, echo=FALSE}
plyr::join_all(list(mod5.nlme.VC %>% dplyr::select(grpA, grpB, Variance) %>% rename(nlme=Variance),
                    mod5.nlme.VC %>% dplyr::select(grpA, grpB) %>% mutate(lme4=NA),
                    mod5.glmm.VC %>% mutate(grpA = str_remove(grpA, "date"),
                                            grpB = str_remove(grpB, "densf")) %>% 
                      dplyr::select(grpA, grpB, estimate) %>% rename(glmmTMB=estimate),
                    mod5.somm.VC %>% mutate(grp = str_replace(grp, ":units", "")) %>% 
                     dplyr::select(grp, VarComp) %>% separate(grp, sep="\\:", into=c("grpA","grpB")) %>%
                     rename(sommer=VarComp),
                    mod5.sas.VC %>% dplyr::select(grpA, grpB, Estimate) %>% rename(SAS=Estimate)), 
               by=c("grpA", "grpB"), type="left") %>% 
  arrange(grpA, grpB) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```

## AIC

```{r, echo=FALSE}
plyr::join_all(list(AIC.nlme %>% dplyr::select(Modnames, AICc) %>% rename(nlme=AICc),
                    AIC.lme4 %>% dplyr::select(Modnames, AICc) %>% rename(lme4=AICc),
                    AIC.glmm %>% dplyr::select(Modnames, AICc) %>% rename(glmmTMB=AICc),
                    AIC.somm %>% dplyr::select(Modnames, AIC)  %>% rename(sommer=AIC),
                    AIC.sas  %>% dplyr::select(Modnames, AICC) %>% rename(SAS=AICC)
                    ), by="Modnames", type="left") %>% 
  mutate_if(is.double, round, 3) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("bordered", "hover", "condensed", "responsive"), 
                full_width = FALSE)
```


# More on this
[Similar chapter on Michael Clark's Website "Mixed Models with R"](https://m-clark.github.io/mixed-models-with-R/extensions.html#heterogeneous-variance){target="_blank"}

["SAS and R" blog post](http://sas-and-r.blogspot.com/2014/02/example-20143-allow-different-variances.html?m=1){target="_blank"}