---
title: "Mixed Models for Agriculture in R"
output: 
  html_document:
    includes:
      after_body: footer.html		
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
# packages for better formatting tables for html output
library(kableExtra); library(formattable)
library(dplyr); library(tibble); library(tidyr)
```

# Hi there!

This is a project under constrcution. Our aim is to provide a cookbook with mixed model analyses of typical examples in life sciences (focus on agriculture/biology) and compare the possibilities or rather limitations of the R-packages

* [`nlme`](https://CRAN.R-project.org/package=nlme){target="_blank"}
* [`lme4`](https://CRAN.R-project.org/package=lme4){target="_blank"}
* [`glmmTMB`](https://CRAN.R-project.org/package=glmmTMB){target="_blank"}
* [`sommer`](https://CRAN.R-project.org/package=sommer){target="_blank"}

to each other, but also to SAS' [`PROC MIXED`](https://documentation.sas.com/?docsetId=statug&docsetTarget=statug_mixed_syntax.htm&docsetVersion=14.3&locale=en){target="_blank"}.

<br/>

# Structure {.tabset .tabset-fade .tabset-pills}
Find examples/chapters by navigating the menu at the top of the page. Each example focuses on a certain aspect of mixed modelling. Within each example, switch between tabs with codes from the different packages via these buttons:

## nlme
## lme4
## glmmTMB
## sommer
## SAS

  <br/><br/>  

# Table of Contents
```{r echo=FALSE, message=FALSE, warning=FALSE}
contents <- rbind(c("Heterogeneous error variance", "yes", "no", "kind of", "in progress", "in progress")) %>% data.frame
names(contents) <- c("Example", "nlme", "lme4", "glmmTMB", "sommer", "SAS")

contents %>% 
  mutate(Example = cell_spec(Example, bold=TRUE)) %>% 
  mutate_at(vars(nlme:SAS), 
            ~cell_spec(., 
                       color=case_when(.=="in progress" ~ "orange", 
                                       TRUE ~ "white"),
                       background=case_when(.=="yes" ~ "green",
                                            .=="kind of" ~ "orange",
                                            .=="no" ~ "red",
                                            TRUE ~ "white"))) %>%
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                position = "left",
                full_width = FALSE) %>% 
  add_header_above(c(" ", "Possible in"=5))
```






